Search for Spatially Extended Sources Using Two Years of Flight Data

### 2011-04-11 extended_catalog/paper/extended_catalog.tex
adding new figure

\documentclass[12pt]{article}

@@

\abstract{
We present a new method for analyzing spatially 
extended sources with the Fermi Large Area
Telescope (LAT), the primary science instrument
on the {\em Fermi Gamma-ray Space Telescope (Fermi)}.
We provide a series of Monte Carlo studies
that were done to validate this tool.
We then apply this tool to test all sources in
the 2FGL Two Year Catalog for extension
and report on the discovery of new extended sources.\cite{2FGL}
}

@@

@@

Performing morphological studies of sources in the GeV energy range using
the Fermi LAT is challenging primarily because of the significantly energy
dependent point spread function and because of systematic errors in the
galactic diffuse emission. The Fermi PSF varies from over $XXXXX^\circ$
at 100MeV to $XXXXX^\circ$ at 100GeV and so the higher energy photons are
significantly more important for studying the shape of sources. One could
ideally look just at the highest energy photons, but the trade off is that
for typical sources many more source events are expected at low energy.
For this reason, it is especially important to use as much of the data
as possible.

@@

Instead, an alternate approach was developed for performing
a morphological analysis. An alternate maximum likelihood fitting
package called {\em pointlike} has been developed by the Fermi
collaboration and we have expanded its functionality to fully support
analyzing spatially extended sources. {\em pointlike} is most thoroughly
described in Matthew Kerr's Ph.D Thesis\cite{Matthew_kerr_phd_thesis}
but the important aspects will be summarized below. What differentiates
{\em pointlike} most significantly from {\em gtlike} is its emphasis
on speed by introducing approximations into the calculation of the
likelihood function while at the same time trying to reduce any
numerical errors that may be introduced.  Like binned {\em gtlike},
{\em pointlike} bins the sky both in position and energy.  Unlike {\em
gtlike}, pointlike relies on the {\em healpix} representation of spatial
bins\cite{healpix_paper} and scales the size of each spatial bin to be
smaller than the point spread function. This is convenient because it no
significant information is lost due to binning at all energies and it is
always a good approximation that the model predicted counts are uniform
across a spatial bin. Furthermore, {\em pointlike} uses a sparse matrix
representation of the spatial bins where only the bins with counts in
them are looped over. The only downside to this is that the integral of
the model predicted counts must be independently calculated.  Effectivly,
{\em pointlike} efficiently interpolates from a totally binned analysis
at low energy (where the resolution is poor and there are many counts)
to an unbinned analysis at high energy (where the resolution is very good
but there are few counts). This approach provides significant savings
in time without introducing large numerical error.

% PseudoHypothesis

\section{Validation}

One can define a parameter $\text{TS}_\text{ext}$ as twice the increase
in log likelihood going from fitting a source with a point hypothesis to fitting
it as an extended hypothesis. So $\text{TS}_\text{ext}=2\times(LL_\text{ext}-LL_\text{ps})$.
In going from fitting a point source to fitting a simple extended source such as
a disk, one degree of freedom, the source width, is added. According to
Wilk's Theorem, the distribution of $\text{TS}_\text{ext}$ for no signal (a true
point source) should follow a $\chi^2$ distribution with one degree of freedom\cite{Wilks_Theorem}.
On the other hand, since the null hypothesis exists on the edge of parameter space,
and so so Wilk's Theorem does not hold\cite{Warnings about Wilk's Thorem}.
One might expect the actual distribution to be a half $\chi^2$ distribution with a delta
function at 0. This was the distribution found to be true for source finding of
gamma ray sources.\cite{Mattox_et_All_Paper}. A Monte Carlo study has been performed to
show how well we can believe the statistical significance.

@@

\section{Results}

>>>>

\documentclass[12pt]{article}
\usepackage{fullpage}

@@

\abstract{
We present a new method for analyzing spatially extended sources with
the Fermi Large Area Telescope (LAT), the primary science instrument
on the {\em Fermi Gamma-ray Space Telescope (Fermi)}.  We provide a
series of Monte Carlo studies that were done to validate this tool.
We then apply this tool to test all sources in the 2FGL Two Year Catalog
for extension.\cite{2FGL} We then report on the discovery of XXX new
extended sources.
}

@@

The Large Area Telescope (LAT) on the Fermi Gamma-Ray Space Telescope
has been surveying the Gamma-ray sky since August 2008. The LAT has been
been essential in studying a variety of astrophysical sources including
Active Galactic Nuclie, Pulsars, Supernova Remnants. 
One of the important data productes

Using 11 months of data, the LAT produced the best-resolved survey of the sky
in the 100 MeV to 100 GeV energy range\cite{2FGL Paper}. The

@@

Performing morphological studies of sources in the GeV energy range using
the Fermi LAT is challenging primarily because of the significantly energy
dependent point spread function (PSF) and because of systematic errors in the
galactic diffuse emission. The Fermi PSF varies from over $XXXXX^\circ$
at 100MeV to $XXXXX^\circ$ at 100GeV and so the higher energy photons are
significantly more important for studying the shape of sources. One could
ideally look just at the highest energy photons, but the trade off is that
for typical sources many more source events are expected at low energy.
For this reason, it is especially important to use as much of the data
as possible.

@@

Instead, an alternate approach was developed for performing
a morphological analysis. An alternate maximum likelihood fitting
package called {\em pointlike} has been developed by the Fermi
collaboration and we have expanded its functionality to fully support
analyzing spatially extended sources. {\em pointlike} is most thoroughly
described in Matthew Kerr's Ph.D Thesis\cite{Matthew_kerr_phd_thesis}
but the important aspects will be summarized below. What differentiates
{\em pointlike} most significantly from {\em gtlike} is its emphasis
on speed by introducing approximations into the calculation of the
likelihood function while at the same time trying to reduce any
numerical errors that may be introduced.  Like binned {\em gtlike},
{\em pointlike} bins the sky both in position and energy.  Unlike {\em
gtlike}, pointlike relies on the {\em healpix} representation of spatial
bins\cite{healpix_paper} and scales the size of each spatial bin to be
smaller than the PSF. This is convenient because it no
significant information is lost due to binning at all energies and it is
always a good approximation that the model predicted counts are uniform
across a spatial bin. Furthermore, {\em pointlike} uses a sparse matrix
representation of the spatial bins where only the bins with counts in them
are looped over. The only downside to this is that the integral of the
model predicted counts must be independently calculated.  Effectively,
{\em pointlike} efficiently interpolates from a totally binned analysis
at low energy (where the resolution is poor and there are many counts)
to an unbinned analysis at high energy (where the resolution is very
good but there are few counts). This approach provides a significant
savings in time while introducing only small numerical error.

We have modified {\em pointlike} to support fitting spatially extended
sources.  To do this, we introduced a new class of diffuse sources where
the spatial and spectral information completely decouple and where
the spatial shape is parameters by simple objects (Disks, Gaussian).
The code consistently convolves the extended source shape with the point
spread function (as a function of energy) and integrates the spectrum over energy
to computed model predicted counts.

Finally, {\em pointlike} allows the spatial parameters of an extended
source to be simulataneoulsy fit. {\em pointlike} uses the Minuit fitting
library to perform the maximum likelihood maximization of the position
and extension of an extended source\cite{Minuit reference}.  At each
step in the fit of the shape, the spectral parameters of the entire sky
model are reoptimized. To avoid projection effects, what is directly
fit is not the longitude and latitude of the source but instead
the displacement in a rotated reference frame. The covariance matrix is
then computed and inverted to estimate the statistical errors on the
position and extension.

{\em pointlike} optimizes the analysis of extended sources in two ways.
First, for extended sources which are radially symmetric, the convolution
of the source shape with our parameterization of the PSF can be performed
semi-analytically. Generally, at a given energy the observed 
photon distribution is
\begin{equation}
  \text{PDF}(\vec r) = \int  \text{PSF}(|\vec r - \vec r'|)I_\text{src}(\vec r') d A' 
\end{equation}
where $I_\text{src}$ is the extended soruce shape and the PSF is parameterized by
\begin{equation}
  \text{PSF}(\vec r) = 
  \frac{1}{2\pi\sigma^2}
  \left(1-\frac{1}{\gamma}\right)
  \left(1+\frac{u}{\gamma}\right)^{-\gamma}
\end{equation}
Here, $\sigma$ and $\gamma$ are parameterize the $PSF$.
If the extended source is radially symmetric, then
$I_\text{src} (\vec r') = I_\text{src} (r')$ and the angular part of the
integral can be done analytically, leaving a simpler one dimensional
integral:
\begin{equation}
  \text{PDF}(u)= \int_0^\infty dv
  I_\text{src}(v) 
  \left(\frac{\gamma-1}{\gamma}\right)
  \left( \frac{\gamma}{\gamma + u + v}\right)^\gamma
  ~_2F_1 \left(\gamma/2,\frac{1+\gamma}{2},1,\frac{4uv}{(\gamma+u+v)^2}\right)
\end{equation}
Not only is this convolution formula faster, but it has the advantage
that it more accurately reduces to the actual PSF shape as the source's
size decreases.  Decreasing the numerical difference between a small extended
source and the PSF is important when searching for smaller extended sources
and when determining the statistical significance of a detection.

There is one final approximation which speeds up the convolution.
Generally, the PSF is a function of the angle $\theta$ of the incident
event measured relative to the normal to the LAT. To avoid repeatedly
convolving the source shape with the PSF at a series of $\theta$ angles,
an effective PSF is fit to the weighted sum of the PSF times the exposure
at different $\theta$ angles.  Since the source convolution has to be
done at each fit iteration, these optimization were essential for making
this tool fast enough for the following analysis. Typically, the entire
process of fitting the extension and estimating the errors for a source
would run on the order of an hour.

\section{Monte Carlo Validation}

One can define a parameter $\text{TS}_\text{ext}$ as twice the increase
in log likelihood going from fitting a source with a point hypothesis to fitting
it as an extended hypothesis. So $\text{TS}_\text{ext}=2\times(LL_\text{ext}-LL_\text{ps})$.
Note that Wilk's Theorem only holds when the likelihood is maximized simultaneoulsy
fitting all of the parameters of model. It is for this reason why {\em pointlike}
goes through such great effort to simultaneously fit the position and extension of
a source.
In going from fitting a point source to fitting a simple extended source such as
a disk, one degree of freedom, the source width, is added. According to
Wilk's Theorem, the distribution of $\text{TS}_\text{ext}$ for no signal (a true
point source) should follow a $\chi^2$ distribution with one degree of freedom\cite{Wilks_Theorem}.
On the other hand, since the null hypothesis exists on the edge of parameter space,
and so so Wilk's Theorem does not hold\cite{Warnings about Wilk's Theorem Paper}.
One might expect the actual distribution to be a half $\chi^2$ distribution with a delta
function at 0. This was the distribution found to be true for source finding of
gamma ray sources.\cite{Mattox_et_All_Paper}. A Monte Carlo study has been performed to
show how well we can believe the statistical significance.

One issue that we found when testing this analysis method is that
despite our best efforts, there was always a small discrepancy between
the likelihood value when fitting the source as a point source and when
fitting it as an extended source with a small extension.  This is to be
expected due to numerical innacuracies introduced by the convolution.
For most cases, this error is insignificant. It would not significanlty
bias any of the fit values or errors. But it would bias the distribution
of $\text{TS}_\text{ext}$ in the null hypothesis. To correct for this,
instead of computing the likelihood in the null hypothesis by fitting
a point source, a new source model was introduced which is an extended
source with its extension fixed to a very small value. $LL_\text{ps}$
is then calculated by refitting this source using the same extended
source fitting code. It was only after introducing this change that we
were able to get broad agreement between simulations and Wilk's theorem.

\includegraphics{k

@@

\section{Test of 1LAC Sources}

\section{Validation of Known Extended Sources}

\section{New Extended Sources}
